---
title: "Lab 6"
author: "Aracely Menjivar"
output: pdf_document
---

#Visualization with the package ggplot2

I highly recommend using the [ggplot cheat sheet](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) as a reference resource. You will see questions that say "Create the best-looking plot". Among other things you may choose to do, remember to label the axes using real English, provide a title and subtitle. You may want to pick a theme and color scheme that you like and keep that constant throughout this lab. The default is fine if you are running short of time.

Load up the `GSSvocab` dataset in package `carData` as `X` and drop all observations with missing measurements. This will be a very hard visualization exercise since there is not a good model for vocab.

```{r}
library(ggplot2)
pacman::p_load(carData)
X = carData::GSSvocab
X = na.omit(X)
summary(X)
```

Briefly summarize the documentation on this dataset. What is the data type of each variable? What do you think is the response variable the collectors of this data had in mind?

Year: Discrete
Gender: Nominal Categorcal
NativeBorn: Nominal Categorical
AgeGroup: Ordinal Categorical
EducGroup: Nominal Categorical
Vocab: Discrete
Age: Continuous
Educ: Continuous

Create two different plots and identify the best-looking plot you can to examine the `age` variable. Save the best looking plot as an appropriately-named PDF.

```{r}
ggplot(X, aes(x=age))+
  geom_histogram(bins=50)
age_Plot = ggplot(X, aes(x=age))+
  geom_bar(kernel  = "guassian")
age_Plot
```

Create two different plots and identify the best looking plot you can to examine the `vocab` variable. Save the best looking plot as an appropriately-named PDF.

```{r}
factor(X$vocab)
ggplot(X, aes(x=vocab))+
  geom_dotplot()
vocab_Plot = ggplot(X, aes(x=vocab))+
  geom_bar()
vocab_Plot
```

Create the best-looking plot you can to examine the `ageGroup` variable by `gender`. Does there appear to be an association? There are many ways to do this.

From the plot below it seems like most people in this dataset are either relatively young or old.

```{r}
ggplot(X)+
  geom_count(aes(x=ageGroup, y=gender))
```

Create the best-looking plot you can to examine the `vocab` variable by `age`. Does there appear to be an association?

The association from the plot below seems to be that there are adults who are either good at vocab or not.

```{r}
X$vocab = factor(X$vocab)
vocabAge_Plot = ggplot(X) +
  geom_boxplot(aes(x=vocab, y=age))
vocabAge_Plot
```

Add an estimate of $f(x)$ using the smoothing geometry to the previous plot. Does there appear to be an association now?

Yes, middle aged people perform the best.We peek vocabulary knowledge in the middle of lifespan.


```{r}
X$vocab = as.numeric(X$vocab)
smoothAgeVocab_Plot = ggplot(X, aes(x=age, y=vocab)) +
  geom_smooth()
smoothAgeVocab_Plot
```

Using the plot from the previous question, create the best looking plot overloading with variable `gender`. Does there appear to be an interaction of `gender` and `age`?

Female and Male seem to have the same amount of vocab knowledge.

```{r}
overloadAgeVocab_Plot = ggplot(X, aes(x=age, y=vocab))+
  geom_jitter(aes(col=gender))+
  geom_smooth()
overloadAgeVocab_Plot
```


Using the plot from the previous question, create the best looking plot overloading with variable `nativeBorn`. Does there appear to be an interaction of `nativeBorn` and `age`?

you tend to not forget your native language as quickly as secondary languages.

```{r}
overloadAgeVocab_Plot2 = ggplot(X, aes(x=age, y=vocab))+
  geom_jitter(aes(col=nativeBorn))+
  geom_smooth()
overloadAgeVocab_Plot2
```

Create two different plots and identify the best-looking plot you can to examine the `vocab` variable by `educGroup`. Does there appear to be an association?

Clearly the highest educated groups are most represented in the higher vocab scores. It follows as education decreases so does vocab score.

```{r}
ggplot(X, aes(x=educGroup, y=vocab))+
  geom_boxplot()
densityEduVocab_Plot = ggplot(X, aes(x=vocab))+
  geom_density(aes(fill=educGroup), adjust=1)
densityEduVocab_Plot
```

Using the best-looking plot from the previous question, create the best looking overloading with variable `gender`. Does there appear to be an interaction of `gender` and `educGroup`?

If years 12 and less, females tend to know more vocabulary. During 13-15 years, female and male tend to be
the same.

```{r}
OverloadGenderEducGroupPlot = ggplot(X, aes(x=educGroup,y=vocab))+
  geom_boxplot(aes(col=gender))
OverloadGenderEducGroupPlot
```

Using facets, examine the relationship between `vocab` and `ageGroup`. You can drop year level `(Other)`. Are we getting dumber?

there is a lot less variation between high and lower performers, so we might be getting dumber.

```{r}
ggplot(GSSvocab) +
aes(x = age, y=vocab) +
geom_jitter(aes(col = gender)) +
geom_smooth(col = "darkseagreen1")
```